Inferring cosmology from gravitational waves using non-parametric detector-frame mass distribution

### 2024-07-17 src/tex/ms.tex
[showyourwork] overleaf sync

% Section outline
In Sec.~\ref{sec:method}, we describe our modified method and the testing procedure.
In Sec.~\ref{sec:simulation}, we present the analysis setup and results of the simulation study.
In Sec.~\ref{sec:real_data}, we apply our method to the real data from the \ac{LVK}.
In Sec.~\ref{sec:discussion}, we compare our results with other methods and discuss future directions.
Finally, we conclude in Sec.~\ref{sec:conclusion} with a summary.
\todo{Rewrite outline}

@@

\begin{table}
    \label{tab:notation}
    \caption{Notation}
    \begin{ruledtabular}
        \begin{tabular}{c>{\raggedright\arraybackslash}p{0.8\linewidth}}
            Notation & Description \\
            \hline
            $m_1$ & Primary mass of the \ac{GW} sources \\
            $z$ & Redshift of the \ac{GW} sources \\
            $d_L$ & Luminosity distance of the \ac{GW} sources \\
            $m^z_1$ & Primary mass of the \ac{BBH} sources in the detector frame \\
            $\mathbf{Y}$ & Set of posterior samples of the \ac{GW} events \\
            $\mathbf{\Theta}$ & Set of hyperparameters of the \ac{(H)DPGMM} \\
            $p(m^z_1|\mathbf{\Theta})$ & Detector-frame observed mass population distribution reconstructed with \ac{(H)DPGMM} \\
            $\Lambda$ & Source-frame population model \\
            $\Omega$ & Cosmological model \\
            $p(m_1|\Lambda)$ & Source-frame mass population distribution assuming the population model $\Lambda$ \\
            $p(m^z_1|\Lambda, \Omega)$ & Detector-frame mass population distribution assuming the population model $\Lambda$ and cosmological model $\Omega$ \\
            $\mathrm{det}$ & detectability of the \ac{GW} events \\
        \end{tabular}
        \end{ruledtabular}
\end{table}

@@

We use a \ac{(H)DPGMM} developed in \citet{Rinaldi:2021bhm} as a non-parametric model to reconstruct the detector-frame observed mass population distribution $p(m^z_1|\mathbf{\Theta})$.
The \ac{(H)DPGMM} is a hierarchy of Gaussian mixture models with a Dirichlet process prior, which allows the model to be flexible to reconstruct arbitrary population distribution from posterior samples.
To perform the reconstruction, we use the package \textsc{figaro}\footnote{\textsc{figaro} is publicly available at \url{https://github.com/sterinaldi/FIGARO} and via \texttt{pip}.} \citep{Rinaldi:2024eep}.
Performing the reconstruction once will give a distribution $p(m^z_1|\Theta_i)$ that is consistent with $\mathbf{Y}$ with a given set of hyperparameters $\Theta_i$.
To account for the uncertainty in the reconstruction, we need to perform the reconstruction multiple times to obtain a set of distributions $p(m^z_1|\mathbf{\Theta})$, where $\mathbf{\Theta}$ is a set of $\Theta_i$.
The uncertainty of the reconstruction can then be quantified by the spread of the distributions.
It is important to note that the reconstruction is only valid in the region where the samples are present.
In Fig.~XXX \todo{Add figure}, we show an example of the reconstruction obtained with \textsc{figaro}.

@@

With different $\Omega$, the resulting $p(m^z_1|\Lambda, \Omega, \mathrm{det})$ will be different.
By considering a continuous set of $\Omega$ and  $\Lambda$ of the population model, we can obtain a set of $p(m^z_1|\mathbf{\Lambda}, \mathbf{\Omega}, \mathrm{det})$, which is a set of continuous detector-frame observed population models for each combination of $\Omega_j$ and $\Lambda_j$, where $\mathbf{\Lambda} = \{\Lambda_j\}$ and $\mathbf{\Omega} = \{\Omega_j\}$.
In Fig.~XXX \todo{Add figure}, we show an example of the transformation.

@@

For each reconstructed distribution $p(m^z_1|\Theta_i)$, we find the optimized $\Lambda_j$ and $\Omega_j$ by minimizing $d_\mathrm{JS}$.
For the minimization, we use the \textsc{scipy} \citep{2020SciPy-NMeth} implementation of the \todo{XXX} algorithm. \todo{Add reference for the algorithm}
This procedure gives us a set of optimized $\Lambda_i$ and $\Omega_j$ for each $\Theta_i$.
Note that we first calculate the median of the detector-frame primary mass posterior samples of each \ac{GW} event, then we calculate $d_\mathrm{JS}$ in the range from the minimum to the maximum of the medians.
This is because the reconstruction is only valid in the region where observations are present.

The resulting distribution of the optimized parameters is not a posterior distribution in the Bayesian sense, as we do not consider the likelihood of the observed data given the population model.
Instead, it is a distribution of the optimized parameters that are consistent with the observed data, distributed according to the uncertainty in the reconstruction of the population.
That is, the uncertainty in the comparison of the two populations is not considered in the resulting distribution.
By considering this optimization procedure only, we can estimate the optimized parameters without performing computationally expensive likelihood calculations.
In Fig.~XXX \todo{Add figure}, we show an example of the fitting procedure.

@@

\section{Simulation}
\label{sec:simulation}

% $H_0$ only

@@

% Population model from simulation

>>>>

% Section outline
In Sec.~\ref{sec:method}, we describe our modified method and the testing procedure.
In Sec.~\ref{sec:mock_data}, we present the analysis setup and results of the mock data study.
In Sec.~\ref{sec:real_data}, we apply our method to the real data from the \ac{LVK}.
In Sec.~\ref{sec:discussion}, we compare our results with other methods and discuss future directions.
Finally, we conclude in Sec.~\ref{sec:conclusion} with a summary.
\todo{Rewrite outline}

@@

\begin{table}
    \label{tab:notation}
    \caption{Notation}
    \begin{ruledtabular}
        \begin{tabular}{c>{\raggedright\arraybackslash}p{0.8\linewidth}}
            Notation & Description \\
            \hline
            $m_1$ & Primary mass of the \ac{GW} sources \\
            $z$ & Redshift of the \ac{GW} sources \\
            $d_L$ & Luminosity distance of the \ac{GW} sources \\
            $m^z_1$ & Primary mass of the \ac{BBH} sources in the detector frame \\
            $Y_t$ & source-frame primary mass posterior samples of the t-th \ac{GW} event \\
            $\mathbf{Y}$ & Set of source-frame primary mass posterior samples of all \ac{GW} events \\
            $\mathbf{\Theta}$ & Set of hyperparameters of the \ac{(H)DPGMM} \\
            $p(m^z_1|\mathbf{\Theta})$ & Detector-frame observed mass population distribution reconstructed with \ac{(H)DPGMM} \\
            $\Lambda$ & Source-frame population model \\
            $\Omega$ & Cosmological model \\
            $p(m_1|\Lambda)$ & Source-frame mass population distribution assuming the population model $\Lambda$ \\
            $p(m^z_1|\Lambda, \Omega)$ & Detector-frame mass population distribution assuming the population model $\Lambda$ and cosmological model $\Omega$ \\
            $\mathrm{det}$ & detectability of the \ac{GW} events \\
        \end{tabular}
        \end{ruledtabular}
\end{table}

@@

We use a \ac{(H)DPGMM} developed in \citet{Rinaldi:2021bhm} as a non-parametric model to reconstruct the detector-frame observed mass population distribution $p(m^z_1|\mathbf{\Theta})$.
The \ac{(H)DPGMM} is a hierarchy of Gaussian mixture models with a Dirichlet process prior, which allows the model to be flexible to reconstruct arbitrary population distribution from posterior samples.
To perform the reconstruction, we use the package \textsc{figaro}\footnote{\textsc{figaro} is publicly available at \url{https://github.com/sterinaldi/FIGARO} and via \texttt{pip}.} \citep{Rinaldi:2024eep}.
Performing the reconstruction once will give a distribution $p(m^z_1|\Theta_i)$ that is consistent with $\mathbf{Y}$ with a given set of hyperparameters $\Theta_i$.
To account for the uncertainty in the reconstruction, we need to perform the reconstruction multiple times to obtain a set of distributions $p(m^z_1|\mathbf{\Theta})$, where $\mathbf{\Theta}$ is a set of $\Theta_i$.
The uncertainty of the reconstruction can then be quantified by the spread of the distributions.
It is important to note that the reconstruction is only valid in the region where the samples are present.
In Fig.~\ref{fig:simulation_reconstruction}, we show an example of the reconstruction obtained with \textsc{figaro}.

@@

With different $\Omega$, the resulting $p(m^z_1|\Lambda, \Omega, \mathrm{det})$ will be different.
By considering a continuous set of $\Omega$ and  $\Lambda$ of the population model, we can obtain a set of $p(m^z_1|\mathbf{\Lambda}, \mathbf{\Omega}, \mathrm{det})$, which is a set of continuous detector-frame observed population models for each combination of $\Omega_j$ and $\Lambda_j$, where $\mathbf{\Lambda} = \{\Lambda_j\}$ and $\mathbf{\Omega} = \{\Omega_j\}$.
In Fig.~\ref{fig:simulation_transformation}, we show an example of the transformation.

@@

For each reconstructed distribution $p(m^z_1|\Theta_i)$, we find the optimized $\Lambda_j$ and $\Omega_j$ by minimizing $d_\mathrm{JS}$.
For the minimization, we use the \textsc{scipy} \citep{2020SciPy-NMeth} implementation of the \todo{XXX} algorithm. \todo{Add reference for the algorithm}
This procedure gives us a set of optimized $\Lambda_i$ and $\Omega_j$ for each $\Theta_i$.
Note that we first calculate the median of the detector-frame primary mass posterior samples of each \ac{GW} event, then we calculate $d_\mathrm{JS}$ in the range from the minimum to the maximum of the medians.
This is because the reconstruction is only valid in the region where observations are present.
In Fig.~\ref{fig:simulation_comparison}, we show an illustration of the comparison between the two populations.

The resulting distribution of the optimized parameters is not a posterior distribution in the Bayesian sense, as we do not consider the likelihood of the observed data given the population model.
Instead, it is a distribution of the optimized parameters that are consistent with the observed data, distributed according to the uncertainty in the reconstruction of the population.
That is, the uncertainty in the comparison of the two populations is not considered in the resulting distribution.
By considering this optimization procedure only, we can estimate the optimized parameters without performing computationally expensive likelihood calculations.

@@

\section{Mock data}
\label{sec:mock_data}

% $H_0$ only
\begin{figure}[h]
    \script{plot_simulation_reconstruction.py}
    \includegraphics[width=\linewidth]{figures/simulation_reconstruction.pdf}
    \caption{
        \todo{Add caption}
    }
    \label{fig:simulation_reconstruction}
\end{figure}

\begin{figure}[h]
    \script{plot_simulation_transformation.py}
    \includegraphics[width=\linewidth]{figures/simulation_transformation.pdf}
    \caption{
        \todo{Add caption}
    }
    \label{fig:simulation_transformation}
\end{figure}

\begin{figure}[h]
    \script{plot_simulation_comparison.py}
    \includegraphics[width=\linewidth]{figures/simulation_comparison.pdf}
    \caption{
        \todo{Add caption}
    }
    \label{fig:simulation_comparison}
\end{figure}

\begin{figure}[h]
    \script{plot_simulation_result_H0.py}
    \includegraphics[width=\linewidth]{figures/simulation_result_H0.pdf}
    \caption{
        \todo{Add caption}
    }
    \label{fig:simulation_result_H0}
\end{figure}

@@
